Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-11 talk/talk.tex
Include Greenland cross section.  Edits.

\begin{itemize}
\item solve one step of backward Euler for $s,\bu,p$ at new time $t^\ell$
    \begin{itemize}
    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$ in evolving SKE, and write $s$ for $s^\ell$
    \end{itemize}
\item again get an NCP coupled to a Stokes problem:

@@

    \begin{itemize}
    \item \alert{similar \dots mostly suffices to consider steady IGP}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Brinkerhoff et al (2017) (\emph{upwinding, semi-coupled})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
    \item neither scalable (\emph{single level}) nor unconditionally stable
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\medskip
    \item \emph{my goal}: unconditionally-stable implicit with Stokes dynamics
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{center}
\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.15\textwidth]{figs/elephant.png}
\end{center}

@@

    \begin{itemize}
    \item $\Phi$ is diffusive in the large!
    \item $\Phi$ is degenerate at the ice margin!
    \end{itemize}
\item yet common to say $\Phi$ is hyperbolic because SKE is transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item $P_1$ because low regularity at free boundary
    \item uses numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$
    \end{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$ 
        \begin{itemize}
        \item[{\color{black} $\circ$}] optimal means work is $O(m)$ \dots requires multigrid
        \end{itemize}
    \item<2> \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\begin{itemize}
\item solve one step of backward Euler for $s,\bu,p$ at new time $t^\ell$
    \begin{itemize}
    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$ in evolving SKE, and write $s$ for $s^\ell$
    \end{itemize}
\item at each time step, solve NCP coupled to Stokes:

@@

    \begin{itemize}
    \item very similar to steady IGP
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\medskip
    \item \emph{my goals}: Stokes, unconditionally-stable steps, scalable
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{center}
\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}
\end{center}

@@

    \begin{itemize}
    \item $\Phi$ is diffusive in the large!
    \item $\Phi$ is degenerate at the ice margin!
    \end{itemize}
\item yet common to say $\Phi$ is hyperbolic because SKE is equivalent to a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item $P_1$ because of low regularity at free boundary
    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ using numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
    \end{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} $O(m)$ work over $m$ degrees of freedom in $\mathcal{V}^h$
        \begin{itemize}
        \item[{\color{black} $\circ$}] requires multigrid
        \end{itemize}
    \item \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{smoothness and multilevel methods}

\begin{columns}
\column{0.35\textwidth}
    \begin{itemize}
    \small
    \item which solution variable?
    \item choice between surface elevation ($s$) and thickness ($H$)
    \item interior smoothness means $s$ is better for multigrid
        \begin{itemize}
        \item need to \emph{not} have fake ice where $s=b$
        \end{itemize}
    \end{itemize}
\column{0.65\textwidth}
    \includegraphics[width=\textwidth]{figs/giscross.png}
\end{columns}
\end{frame}
